Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-03 abstract.tex
New discussion + minor edits
Aryeh's changes
mostly grammatical/spelling; nontrivial ones indicated by \todoa
Merge with Aryeh's version + some further edits

%!TEX root =  matrix-est.tex
This article provides the first procedure for computing a fully
data-dependent interval that traps the mixing time $\tmix$ of a finite
reversible ergodic Markov chain at a prescribed confidence level.
The interval is computed from a single finite-length sample
path from the Markov chain, and does not require the 
knowledge of any parameters of the chain.
%bounds on the mixing time itself.
This stands in contrast to previous approaches, which either 
only provide point estimates, or require a
reset mechanism, or additional prior knowledge.
%Most notably, 
%the procedure produces a fully empirical confidence
%interval (i.e., it depends only on the sample path and a given
%confidence parameter $\delta$) that is always valid with probability
%$1-\delta$; and 
The width of the interval converges to zero at a
$\sqrt{n}$ rate, where $n$ is the length of the sample path.
Upper and lower bounds are given on the number of samples required to
achieve constant-factor multiplicative accuracy.
\todoc{We may need to add some explanation to connect $\gap$ estimates
to $\tmix$ estimates.}
The lower bounds indicate that, unless further restrictions are placed
on the chain, no procedure can achieve this accuracy level before
seeing each state at least $\Omega(\tmix)$ times on the average.
Future directions of research are identified.
\todoc{Such as: closing the gap. Extension to large state spaces, 
based on additional assumptions. Even if $\gap$ is large, we need many samples.
Maybe we need even stronger assumptions, like the decay rate of eigenvalues.
}
%If $\gap$ denotes the absolute spectral gap of the unknown Markov chain
%and the chain's stationary distribution is minorized by $\pimin$,
%our procedure requires $\tilde{O}( \frac{1}{\pimin\gap^3})$ observations
% to estimate the mixing time up to a constant multiplicative accuracy, 
%while we also establish that any algorithm to achieve the same
%goal would need $\Omega( d\log(d)/\gap + 1/\pimin)$ samples.

>>>>

%!TEX root =  matrix-est.tex
This article provides the first procedure for computing a fully
data-dependent interval that traps the mixing time $\tmix$ of a finite
reversible ergodic Markov chain at a prescribed confidence level.
The interval is computed from a single finite-length sample
path from the Markov chain, and does not require the 
knowledge of any parameters of the chain.
%bounds on the mixing time itself.
This stands in contrast to previous approaches, which either 
only provide point estimates, or require a
reset mechanism, or additional prior knowledge.
%Most notably, 
%the procedure produces a fully empirical confidence
%interval (i.e., it depends only on the sample path and a given
%confidence parameter $\delta$) that is always valid with probability
%$1-\delta$; and 
The width of the interval converges to zero at a
$\sqrt{n}$ rate, where $n$ is the length of the sample path. \todoc{Aryeh prefers: .. at a rate of $\sqrt{n}$. }
Upper and lower bounds are given on the number of samples required to
achieve constant-factor multiplicative accuracy.
\todoc{We may need to add some explanation to connect $\gap$ estimates
to $\tmix$ estimates.}
The lower bounds indicate that, unless further restrictions are placed
on the chain, no procedure can achieve this accuracy level before
seeing each state at least $\Omega(\tmix)$ times on the average.
Finally, future directions of research are identified.
%If $\gap$ denotes the absolute spectral gap of the unknown Markov chain
%and the chain's stationary distribution is minorized by $\pimin$,
%our procedure requires $\tilde{O}( \frac{1}{\pimin\gap^3})$ observations
% to estimate the mixing time up to a constant multiplicative accuracy, 
%while we also establish that any algorithm to achieve the same
%goal would need $\Omega( d\log(d)/\gap + 1/\pimin)$ samples.

### 2015-06-03 intro-nips.tex
Merge with Aryeh's version + some further edits

This work
tackles the challenge of constructing
\todoa{was:
considers the construction of}
fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The \emph{mixing time} $\tmix$ of the Markov chain is the
number of time steps required
%time that it
%takes
for the chain to
be within a fixed threshold of
%(approximately) reach
its stationary
distribution:
\todoa{slightly changed wording}
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that
succeeds with probability $1-\delta$
in trapping
the value of the
mixing time $\tmix$.
\todoa{changed wording}

This problem is motivated by the numerous scientific applications and
machine learning tasks in which
the quantity of interest is
%one is interested in estimating
the
mean $\vpi(f) = \sum_i \pi_i f(i)$
for some function $f$ of the states of
a Markov chain.
\todoa{changed wording}
This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
paradigm~\cite{liu2001monte}, but the problem also arises in
performance prediction involving time-correlated data, as is common in
reinforcement learning~\cite{sutton98}.
Observable bounds on mixing times are useful in the design and
diagnostics of these methods; they yield effective approaches to
assessing the estimation quality, even when \emph{a priori} knowledge
of the mixing time or correlation structure is unavailable.

\paragraph{Main results.}
We develop the first procedure for constructing non-trivial
and fully empirical confidence intervals for Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with absolute
spectral gap $\gap$ and stationary distribution minorized by $\pimin$. 
As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
\begin{equation}
  \label{eq:mixing-time-bound}
  \Parens{\frac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \frac1{\gap} \ln \frac4{\pimin}
  ,
\end{equation}
%$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,
and hence it suffices to estimate $\gap$ and $\pimin$. 
%these quantities $\gap$ and $\pimin$ are well-known to govern
%mixing time.
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{sec:rates-upper}, we give a point-estimator for $\gap$,
    and prove in \cref{thm:err} that it achieves multiplicative
    accuracy from \emph{a single sample path} of length
    $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide a point-estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    This establishes the feasibility of \emph{estimating} the mixing
    time in this setting.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.
    We also discuss the importance of reversibility, and some possible
    extensions to non-reversible chains.

@@

    \todoc{Is it true from the construction of \cref{thm:combined}
    that if we had a better point estimator
    then we would automatically get an interval estimator
    with the same accuracy?
    This seems to be true.
    This would mean that to close the gap
    one should really work on coming up with better point estimators.
    }

@@

%with additional access 
when allowed more flexible

@@

\todoa{slightly changed wording}

>>>>

This work
tackles the challenge of constructing
fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The \emph{mixing time} $\tmix$ of the Markov chain is the
number of time steps required
%time that it
%takes
for the chain to
be within a fixed threshold of
%(approximately) reach
its stationary
distribution:
\todoa{slightly changed wording}
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that
succeeds with probability $1-\delta$
in trapping
the value of the
mixing time $\tmix$.
\todoa{changed wording}

This problem is motivated by the numerous scientific applications and
machine learning tasks in which
the quantity of interest is
%one is interested in estimating
the
mean $\vpi(f) = \sum_i \pi_i f(i)$
for some function $f$ of the states of
a Markov chain.
\todoa{changed wording} \todoc{I am not sure that the new wording is better. It
fails to express that there is an estimation problem. "Quantity of interest"
is way to vague for me.} 
This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
paradigm~\cite{liu2001monte}, but the problem also arises in
performance prediction involving time-correlated data, as is common in
reinforcement learning~\cite{sutton98}.
Observable bounds on mixing times are useful in the design and
diagnostics of these methods; they yield effective approaches to
assessing the estimation quality, even when \emph{a priori} knowledge
of the mixing time or correlation structure is unavailable.

\paragraph{Main results.}
We develop the first procedure for constructing non-trivial
and fully empirical confidence intervals for Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with absolute
spectral gap $\gap$ and stationary distribution minorized by $\pimin$. 
As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
\begin{equation}
  \label{eq:mixing-time-bound}
  \Parens{\frac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \frac1{\gap} \ln \frac4{\pimin}
  ,
\end{equation}
%$(\frac{1}\gap  - 1)\ln(2)  \le \tmix \le \frac{1}\gap \ln \frac4\pimin$,
and hence it suffices to estimate $\gap$ and $\pimin$. 
%these quantities $\gap$ and $\pimin$ are well-known to govern
%mixing time.
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{sec:rates-upper}, we give a point-estimator for $\gap$,
    and prove in \cref{thm:err} that it achieves multiplicative
    accuracy from \emph{a single sample path} of length
    $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide a point-estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    This establishes the feasibility of \emph{estimating} the mixing
    time in this setting.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.
    We also discuss the importance of reversibility, and some possible
    extensions to non-reversible chains.

@@

@@

%with additional access to the Markov chain.
when allowed more flexible access

@@
